On the propriety of restricted Boltzmann machines

### 2016-11-30 writing/draft.Rmd
edits from dr. nordman

First, the models are often unsatisfactory as conceptualizations of how data are generated. Recalling [@fisher1922mathematical], the aim of a statistical model is to represent data in a compact way. Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination]. At issue, RBMs  often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of a data generation process. In addition to such degeneracy, we find that RBMs can easily exhibit types of instability. In practice, this may seen when a single pixel change in an image results in a wildly different classification in an image classification problem. Such model impropriety issues have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep]. We investigate these phenomena for RBMs in Section \ref{explorations} through simulations of small, manageable examples.

Secondly, the fitting of these models is problematic. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable. The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about acheiving convergence [@hinton2010practical; @hinton2006fast]. Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.

@@

For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive]. However, the empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data. As a consequence, we find that any fully principled fitting method based on the likelihood for RBM with enough hidden variables will simply seek to reproduce the (discrete) empirical distribution of a training set. Thus, there is no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way without intentially limiting the flexibility of the fitted model.

@@

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making simple estimation of the model parameter vector problematic. A kind of Gibbs sampling can be tried (due to conditional architecture of the RBM, i.e. visibles given hiddens or vice versa). Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).

@@

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of statistics based on the adjacency matrix of a graph. Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$. This makes sense because if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ and can be expected to *not* be pulled deep into the interior of $C$.

A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$. Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
$$

@@

Considering exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*. Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how RBM models might be expanded to incorporate more and more visibles. When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $\nv$ of visibles $(V_1, \dots, V_\nv)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{\nv}$ as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \label{eqn:elpr}
\end{align}
where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameter vector $\boldsymbol \theta_\nv$.

@@

\begin{defn}[S-unstable RBM]
Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles. A RBM model formulation is \emph{Scheinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $n_C > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > C$ for all $\nv \ge n_C$.

@@

In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability). Recall the applied example of RBM models as a means to classify images. This instability result manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image. Examples of this behavior have been presented in @szegedy2013intriguing for other deep learning models, in which a one pixel change in a test image results in a wildly different classification.

@@

```{r instab, fig.cap="\\label{fig:instab}$\\text{ELPR}(\\boldsymbol \\theta)/\\nv$ for various values of $\\boldsymbol \\theta$ for the tiny example model. (Recall that here $\\nv$ is the number of visible nodes and here is $1$. This quantity is large for large magnitudes of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(Q = cbind(possibles, possibles %*% t(data.matrix(.))))) %>%
  group_by(theta_v1, theta_h1, theta_11, Q.v1) %>%
  do(data.frame(marginal = sum(exp(.$Q.V4)))) %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(max_marg = max(.$marginal),
                min_marg = min(.$marginal))) %>%
  mutate(elpr = log(max_marg/min_marg)) %>%
  ungroup() %>%
  mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11)))) %>%
  ggplot() +
  geom_tile(aes(theta_v1, theta_h1, fill = elpr)) +
  geom_contour(aes(theta_v1, theta_h1, z = elpr), colour = "black") +
  facet_grid(. ~ theta_11, labeller = label_parsed) +
  xlab(expression(theta[v1])) +
  ylab(expression(theta[h1])) +
  scale_fill_gradient(expression(frac(ELPR(theta), V)), low = "yellow", high = "red") +
  theme(aspect.ratio = 1)
```

@@

```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
                mu_v1_dep = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_h1_dep = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
  mutate_each(funs(./gamma), starts_with("mu")) %>%
  select(-gamma) -> exp_dep

@@

Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$ and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of $\theta_{11}$. This is a third indication that parameter vectors of large magnitude lead to model impropriety in a RBM.

@@

In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector that serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$). The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the on-line supplementary materials. Hence while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible due mainly to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$ in a model (\ref{eqn:pmf}) of any realistic size. Ad hoc methods are used instead, which aim to avoid this problem by using stochastic ML approximations that employ a small number of MCMC draws (i.e., contrastive divergence, [@hinton2002training]).

However, computational concerns are not the only issues with fitting a RBM using ML. In addition, a RBM model, with the appropriate choice of parameters and number of hiddens, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the $-1/1$ valued data outcomes). For example, @montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by a RBM with $2^{\nv}/2 - 1$ hidden units. To illustrate, we provide a small example.

@@

To demonstrate, =we  assume $p_{(-1,-1)}>0$ (without loss of generality) in the specified cell probabilities (\ref{eqn:0}) and replace parameters $\theta_{11},\theta_{21}$ with $\Delta_1 \equiv \theta_{11} +\theta_{21}$ and $\Delta_2 \equiv \theta_{11} -\theta_{21}$.  We may then prescribe values of $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2$  so that the model probability ratio
$$
P(V_1=v_1,V_2=v_2|\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)/  P(V_1=-1,V_2=-1| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)
$$
matches the corresponding ratio $p_{(v_1,v_2)}/p_{(-1,-1)}$ over three values of $(v_1,v_2) = (1,-1),(-1,1),(1,1)$.

@@

1. *A "trick" prior.* Here we cancel out normalizing term in the likelihood so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal. The $h_j$ are carried allong in the MCMC sampling from the posterior as latent variables.
    \begin{align*}
    \pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right), \vspace{-.75cm}
    \end{align*}

@@

cbind(M_eff_tmp[1:8,], M_eff_tmp[9:16,]) %>%
  kable(digits = 2, caption = paste0("\\label{tab:m_eff}The effective sample sizes for a chain of length $M = 1000$ of all $16$ possible vector outcomes of visibles. BwTNLV would require at least $", round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1), "$ times as many MCMC iterations to achieve the same amount of effective information about the posterior distribution."))
```

>>>>

First, the models are often unsatisfactory as conceptualizations of how data are generated. Recalling @fisher1922mathematical, the aim of a statistical model is to represent data in a compact way. Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination]. At issue, RBMs  often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of a data generation process. In addition to such degeneracy, we find that RBMs can easily exhibit types of instability. In practice, this may seen when a single pixel change in an image results in a wildly different classification in an image classification problem. Such model impropriety issues have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep]. We investigate these phenomena for RBMs in Section \ref{explorations} through simulations of small, manageable examples.

Secondly, the fitting of these models is problematic. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable. The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about achieving convergence [@hinton2010practical; @hinton2006fast]. Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.

@@

For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive]. However, the empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data. As a consequence, we find that any fully principled fitting method based on the likelihood for RBM with enough hidden variables will simply seek to reproduce the (discrete) empirical distribution of a training set. Thus, there can be no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.

@@

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making simple estimation of the model parameter vector problematic. A kind of Gibbs sampling can be tried (due to conditional architecture of the RBM, i.e. visibles given hiddens or vice verse). Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).

@@

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of statistics based on the adjacency matrix of a graph. Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$, under the model above. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$. This makes sense because if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ and can be expected to *not* be pulled deep into the interior of $C$.

A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$. Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from (\ref{eqn:t}), appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
$$

@@

Considering exponential families of distributions, @schweinberger2011instability introduced a concept of model deficiency related to *instability*. Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how RBM models might be expanded to incorporate more and more visibles. When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $\nv$ of visibles $(V_1, \dots, V_\nv)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{\nv}$ as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \label{eqn:elpr}
\end{align}
where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameter vector $\boldsymbol \theta_\nv$.

@@

\begin{defn}[S-unstable RBM]
Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles. A RBM model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $n_C > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > C$ for all $\nv \ge n_C$.

@@

In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability). Recall the applied example of RBM models as a means to classify images. For data as pixels in an image, the instability result in Proposition \ref{prop:instab} manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image. Examples of this behavior have been presented in @szegedy2013intriguing for other deep learning models, in which a one pixel change in a test image results in a wildly different classification.

@@

```{r instab, fig.cap="\\label{fig:instab}$\\text{ELPR}(\\boldsymbol \\theta)/\\nv$ for various values of $\\boldsymbol \\theta$ for the tiny example model. Recall here $\\nv$ is the number of visible nodes and here is $1$. This quantity is large for large magnitudes of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(Q = cbind(possibles, possibles %*% t(data.matrix(.))))) %>%
  group_by(theta_v1, theta_h1, theta_11, Q.v1) %>%
  do(data.frame(marginal = sum(exp(.$Q.V4)))) %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(max_marg = max(.$marginal),
                min_marg = min(.$marginal))) %>%
  mutate(elpr = log(max_marg/min_marg)) %>%
  ungroup() %>%
  mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11)))) %>%
  ggplot() +
  geom_tile(aes(theta_v1, theta_h1, fill = elpr)) +
  geom_contour(aes(theta_v1, theta_h1, z = elpr), colour = "black") +
  facet_grid(. ~ theta_11, labeller = label_parsed) +
  xlab(expression(theta[v1])) +
  ylab(expression(theta[h1])) +
  scale_fill_gradient(expression(frac(ELPR(theta), V)), low = "yellow", high = "red") +
  theme(aspect.ratio = 1)
```

@@

```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node. As an indicator of uninterpretability, note that differences in expectations increase as the dependence parameter $\theta_{11}$ deviates from zero.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
                mu_v1_dep = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_h1_dep = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
  mutate_each(funs(./gamma), starts_with("mu")) %>%
  select(-gamma) -> exp_dep

@@

Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$ and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of the dependence parameter $\theta_{11}$. This is a third indication that parameter vectors of large magnitude lead to model impropriety in a RBM.

@@

In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector then serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$). The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the on-line supplementary materials. Hence while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible due mainly to the intractability of the normalizing term $\gamma(\boldsymbol \theta)$ in a model (\ref{eqn:pmf}) of any realistic size. Ad hoc methods are used instead, which aim to avoid this problem by using stochastic ML approximations that employ a small number of MCMC draws (i.e., contrastive divergence, [@hinton2002training]).

However, computational concerns are not the only issues with fitting a RBM using ML. In addition, a RBM model, with the appropriate choice of parameters and number of hiddens, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the binary data outcomes). For example, @montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by a RBM with $2^{\nv}/2 - 1$ hidden units. To illustrate, we provide a small example.

@@

To demonstrate, we  assume $p_{(-1,-1)}>0$ (without loss of generality) in the specified cell probabilities (\ref{eqn:0}) and replace parameters $\theta_{11},\theta_{21}$ with $\Delta_1 \equiv \theta_{11} +\theta_{21}$ and $\Delta_2 \equiv \theta_{11} -\theta_{21}$.  We may then prescribe values of $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2$  so that the model probability ratio
$$
P(V_1=v_1,V_2=v_2|\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)/  P(V_1=-1,V_2=-1| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)
$$
matches the corresponding ratio $p_{(v_1,v_2)}/p_{(-1,-1)}$ over three values of $(v_1,v_2) = (1,-1),(-1,1),(1,1)$.

@@

1. *A "trick" prior.* Here we cancel out normalizing term in the likelihood so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal. The $h_j$ are carried along in the MCMC sampling from the posterior as latent variables.
    \begin{align*}
    \pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right), \vspace{-.75cm}
    \end{align*}

@@

cbind(M_eff_tmp[1:8,], M_eff_tmp[9:16,]) %>%
  kable(digits = 2, caption = paste0("\\label{tab:m_eff}The effective sample sizes for a chain of length $M = 1000$ regarding all $16$ probabilities for possible vector outcomes of visibles. BwTNLV would require at least $", round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1), "$ times as many MCMC iterations to achieve the same amount of effective information about the posterior distribution."))
```
